Increased Emergency Medical Response Wait Times: The Impact of Higher Call Volume and More Severe Cases Post-COVID-19

### 2025-06-19 paper/paper.qmd
translated paper code to python

```{r}
#| include: false
#| warning: false
#| message: false
#| 
# Package loading
# install.packages("modeest")
library(tidyverse)
library(dplyr)
library(here)
library(tibble)
library(opendatatoronto)
library(ggplot2)
library(knitr)
library(modeest)

# Read data
raw_data <- read_csv(here("data/raw_data/raw_ambulance_response_data.csv"))
cleaned_data <- read_csv(here("data/analysis_data/cleaned_ambulance_response_data.csv"))
```
\newpage

@@

```{r}
#| label: tbl-raw-data-1
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset
#| echo: false
#| warning: false
#| message: false

head_raw_data <- head(raw_data)
# Note there is a typo in Description within the original data set itself
knitr::kable(head_raw_data[,1:5], "simple",   col.names = c('ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene'))
```

```{r}
#| label: tbl-raw-data-2
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset (cont.)
#| echo: false
#| warning: false
#| message: false

# Table is broken into 2 sections to show all data on a single page
knitr::kable(head_raw_data[,6:9], "simple",   col.names = c( 'Forward_Sortation_Area', 'Field Name', 'Desctiption/Definition', 'Comments/Examples'))
```

@@

```{r}
#| label: tbl-cleaned-data-1
#| tbl-cap: First 6 entires of the cleaned ambulance response data
#| echo: false
#| warning: false
#| message: false

head_cleaned_data <- head(cleaned_data)
# Note there is a type in Description within the original data set itself
knitr::kable(head_cleaned_data[,1:5], "simple",   col.names = c('dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene', 'year'))
```

@@

```{r}
#| label: tbl-cleaned-data-summary
#| tbl-cap: Summary of Incident Data Statistics
#| echo: false
#| warning: false
#| message: false

# calculate statistics
mode_incident_type <- mlv(cleaned_data$incident_type, method = "mfv")
mode_priority_number <- mlv(cleaned_data$priority_number, method = "mfv")
mean_units_at_scene <- mean(cleaned_data$units_arrived_at_scene)

total_dispatches_per_year <- cleaned_data %>%
  group_by(year) %>%
  summarise(total_dispatches = n())
mean_dispatches_per_year <- mean(total_dispatches_per_year$total_dispatches)

# Create a data frame to show values
summary_table <- data.frame(
  Field_Name = c("Mean Dispatches per Year", "Mean Units Arrived at Scene", "Mode Incident Type", "Mode Priority Number"),
  Value = c(mean_dispatches_per_year, mean_units_at_scene, mode_incident_type, mode_priority_number)
)

# Print the table
kable(summary_table)
```

```{r}
#| label: fig-year-vs-dispatches
#| fig-cap: Year vs. Total Number of Dispatches
#| echo: false
#| warning: false
#| message: false

cleaned_data |>
ggplot(aes(x=year))+ 
  xlab("Year")+
  ylab("# of Dispatches")+
  geom_bar(position = "dodge")+ 
  scale_x_continuous(breaks = seq(min(cleaned_data$year), max(cleaned_data$year), by = 1)) + 
  theme_minimal()

@@

```{r}
#| label: fig-year-vs-priority
#| fig-cap: Year vs. Priority Number
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each priority number appears each year
priority_count_per_year <- cleaned_data %>%
  filter(!priority_number %in% c(11,12,13,14)) %>%  # Exclude specific priority numbers
  group_by(year, priority_number) %>%
  summarise(count = n(), .groups = 'drop') %>%  # Summarize the count and drop extra grouping
  arrange(year, priority_number)

# Plot years vs priority number
ggplot(priority_count_per_year, aes(x = year, y = count, color = as.factor(priority_number), group = priority_number)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Change in Priority Number Over the Years",
       x = "Year",
       y = "Count of Priority Number",
       color = "Priority Number") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(priority_count_per_year$year))

@@

```{r}
#| label: fig-year-vs-1-unit-arrived
#| fig-cap: Year vs. Number of Times 1 Unit Arrived at Scene
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each priority number appears each year
units_count_per_year <- cleaned_data %>%
  group_by(year, units_arrived_at_scene) %>%
  summarise(count = n()) %>%
  # Filter out low numbers
  filter(count>100000) %>%
  arrange(year, units_arrived_at_scene)

# Plot years vs priority number
ggplot(units_count_per_year, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Change in Units Dispatched Over the Years",
       x = "Year",
       y = "Units Arrived at Scene",
       color = "Units Arrived at Scene") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(units_count_per_year$year))
```

```{r}
#| label: fig-year-vs-2-units-arrived
#| fig-cap: Year vs. Number of Times 2 Units Arrived at Scene
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each priority number appears each year
units_count_per_year2 <- cleaned_data %>%
  group_by(year, units_arrived_at_scene) %>%
  summarise(count = n()) %>%
  #Filter out low and extremely high numbers
  filter(count<100000 & count>10000) %>%
  arrange(year, units_arrived_at_scene)

# Plot years vs priority number
ggplot(units_count_per_year2, aes(x = year, y = count, color = as.factor(units_arrived_at_scene), group = units_arrived_at_scene)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Change in Units Dispatched Over the Years",
       x = "Year",
       y = "Units Arrived at Scene",
       color = "Units Arrived at Scene") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(units_count_per_year$year))
```

>>>>

```{python}
#| include: false
#| warning: false
#| message: false

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt

# Read data
raw_data = pd.read_csv("../data/raw_data/raw_ambulance_response_data.csv")
cleaned_data = pd.read_csv("../data/analysis_data/cleaned_ambulance_response_data.csv")
```
\newpage

@@

```{python}
#| label: tbl-raw-data-1
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset
#| echo: false
#| warning: false
#| message: false

head_raw_data = raw_data.head()

# Select first 5 columns and rename them
table = head_raw_data.iloc[:, 0:5].copy()
table.columns = ['ID', 'Dispatch_Time', 'Incident_Type', 'Priority_Number', 'Units_Arrived_At_Scene']

# Display the table (Quarto will render this as a formatted table)
table
```

```{python}
#| label: tbl-raw-data-2
#| tbl-cap: First 6 entires of the Paramedic Services Incident Dataset (cont.)
#| echo: false
#| warning: false
#| message: false

table_part2 = head_raw_data.iloc[:, 5:9].copy()
table_part2.columns = ['Forward_Sortation_Area', 'Field Name', 'Desctiption/Definition', 'Comments/Examples']

table_part2
```

@@

```{python}
#| label: tbl-cleaned-data-1
#| tbl-cap: First 6 entires of the cleaned ambulance response data
#| echo: false
#| warning: false
#| message: false

head_cleaned_data = cleaned_data.head()

# Select first 5 columns and rename
table_cleaned = head_cleaned_data.iloc[:, 0:5].copy()
table_cleaned.columns = ['dispatch_time', 'incident_type', 'priority_number', 'units_arrived_at_scene', 'year']

table_cleaned
```

@@

```{python}
#| label: tbl-cleaned-data-summary
#| tbl-cap: Summary of Incident Data Statistics
#| echo: false
#| warning: false
#| message: false

# calculate statistics
# Mode for categorical column (incident_type)
mode_incident_type = cleaned_data["incident_type"].mode().iloc[0]

# Mode for numerical column (priority_number)
mode_priority_number = cleaned_data["priority_number"].mode().iloc[0]

# Mean for numerical column (units_arrived_at_scene)
mean_units_at_scene = cleaned_data["units_arrived_at_scene"].mean()

# Total dispatches per year
total_dispatches_per_year = (
    cleaned_data.groupby("year")
    .size()
    .reset_index(name="total_dispatches")
)
mean_dispatches_per_year = total_dispatches_per_year["total_dispatches"].mean()

# Create a data frame to show values
summary_table = pd.DataFrame({
    "Field_Name": [
        "Mean Dispatches per Year",
        "Mean Units Arrived at Scene",
        "Mode Incident Type",
        "Mode Priority Number"
    ],
    "Value": [
        round(mean_dispatches_per_year, 2),
        round(mean_units_at_scene, 2),
        mode_incident_type,
        mode_priority_number
    ]
})

# Print the table
summary_table
```

```{python}
#| label: fig-year-vs-dispatches
#| fig-cap: Year vs. Total Number of Dispatches
#| echo: false
#| warning: false
#| message: false

# Count number of dispatches per year
dispatch_counts = cleaned_data["year"].value_counts().sort_index()

# Create numeric positions for bars
x_pos = np.arange(len(dispatch_counts))  # e.g., [0, 1, 2, 3]

# Plot
plt.figure(figsize=(8, 5))
plt.bar(x_pos, dispatch_counts.values)

# Add axis labels and ticks
plt.xlabel("Year")
plt.ylabel("# of Dispatches")
plt.xticks(ticks=x_pos, labels=dispatch_counts.index.astype(str), rotation=45, ha='right')

# Add styling
plt.title("")
plt.grid(axis='y', linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| label: fig-year-vs-priority
#| fig-cap: Year vs. Priority Number
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each priority number appears each year
priority_count_per_year = (
    cleaned_data[~cleaned_data["priority_number"].isin([11, 12, 13, 14])]
    .groupby(["year", "priority_number"])
    .size()
    .reset_index(name="count")
    .sort_values(["year", "priority_number"])
)

# Plot years vs priority number
plt.figure(figsize=(10, 6))

for priority in priority_count_per_year["priority_number"].unique():
    df_subset = priority_count_per_year[priority_count_per_year["priority_number"] == priority]
    plt.plot(df_subset["year"], df_subset["count"], label=str(priority), marker='o', linewidth=1.5)

plt.title("Change in Priority Number Over the Years")
plt.xlabel("Year")
plt.ylabel("Count of Priority Number")
plt.xticks(sorted(priority_count_per_year["year"].unique()))
plt.legend(title="Priority Number", title_fontsize=10)
plt.grid(axis='y', linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

```{python}
#| label: fig-year-vs-1-unit-arrived
#| fig-cap: Year vs. Number of Times 1 Unit Arrived at Scene
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each units_arrived_at_scene value appears each year
units_count_per_year = (
    cleaned_data
    .groupby(["year", "units_arrived_at_scene"])
    .size()
    .reset_index(name="count")
)

# Filter out low-frequency combinations
units_count_per_year = units_count_per_year[units_count_per_year["count"] > 100000]

# Sort for consistency
units_count_per_year = units_count_per_year.sort_values(["year", "units_arrived_at_scene"])

# Plot years vs units_arrived_at_scene
plt.figure(figsize=(10, 6))

for units in units_count_per_year["units_arrived_at_scene"].unique():
    df_subset = units_count_per_year[units_count_per_year["units_arrived_at_scene"] == units]
    plt.plot(df_subset["year"], df_subset["count"], label=str(units), marker='o', linewidth=1.5)

plt.title("Change in Units Dispatched Over the Years")
plt.xlabel("Year")
plt.ylabel("Units Arrived at Scene")
plt.xticks(sorted(units_count_per_year["year"].unique()))
plt.legend(title="Units Arrived at Scene", title_fontsize=10)
plt.grid(axis='y', linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-year-vs-2-units-arrived
#| fig-cap: Year vs. Number of Times 2 Units Arrived at Scene
#| echo: false
#| warning: false
#| message: false

# Calculate number of times each units_arrived_at_scene value appears each year
units_count_per_year2 = (
    cleaned_data
    .groupby(["year", "units_arrived_at_scene"])
    .size()
    .reset_index(name="count")
)

# Filter out low and extremely high values
units_count_per_year2 = units_count_per_year2[
    (units_count_per_year2["count"] < 100000) & (units_count_per_year2["count"] > 10000)
]

# Sort for consistent plotting
units_count_per_year2 = units_count_per_year2.sort_values(["year", "units_arrived_at_scene"])

# Plot years vs units_arrived_at_scene
plt.figure(figsize=(10, 6))

for units in units_count_per_year2["units_arrived_at_scene"].unique():
    df_subset = units_count_per_year2[units_count_per_year2["units_arrived_at_scene"] == units]
    plt.plot(df_subset["year"], df_subset["count"], label=str(units), marker='o', linewidth=1.5)

plt.title("Change in Units Dispatched Over the Years")
plt.xlabel("Year")
plt.ylabel("Units Arrived at Scene")
plt.xticks(sorted(units_count_per_year2["year"].unique()))
plt.legend(title="Units Arrived at Scene", title_fontsize=10)
plt.grid(axis='y', linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()
```
